Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-01-31 src/tex/ms.tex
hierarchical

\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        The observed Fourier amplitude of two polarizations of a simulated GW signal with and without birefringence.
        The amplitudes, as GR predicts, are shown in dotted lines with blue and orange colors representing the left-handed and right-handed polarizations.
        The amplitudes with birefringence are shown in solid lines with blue and orange colors representing the left-handed and right-handed polarizations.
        This plot shows the effect of amplitude birefringence with frequency dependence.
    }
    \label{fig:birefringence}
\end{figure}

@@

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby}, that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline waveform.
We depart from the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3} \mi{cite actual zenodo link}, with \textsc{IMRPhenomXPHM} as the reference waveform.
We apply a distance prior uniform in comoving volume \mi{is this right?}, and set the prior on $\kappa$ to a uniform distribution between $-1$ and $1$.

\mi{list here any special considerations for individual events}

@@

Since birefringence is a property of spacetime and we are targeting a propagation effect not intrinsic to any source, we should expect $\kappa$ to take the same value for all signals, whether it vanishes or not.
\mi{caveats?}
Under this assumption, the posterior on $\kappa$ inferred collectively from all events with a uniform prior is simply the product of the individual posteriors,%
\footnote{Had the $\kappa$ prior been nonuniform, then the individual \emph{likelihoods} would need to be multiplied, as opposed to the posteriors.}
\begin{equation}
    p(\kappa \mid \{d_i\})\propto \prod_{i}p(\kappa \mid d_i)\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event.
We use this expression to obtain the primary constraint presented in this work.

@@

% Bayesian Hierarchical Modeling
In order to further characterize our set of measurements, and identify potential outliers, we carry out an additional collective analysis that allows $\kappa$ to take different values for each events.
To do this, we make use of hierarchical Bayesian inference to model the distribuion of $\kappa$'s consistent with the observed data.
Here, instead of assuming all events share the same value of $\kappa$, we assume $\kappa$ from each event is a true value drawn from a Gaussian distribution with mean $\mu$ and standard deviation $\sigma$.
We chose a Gaussian distribution as the underlying distribution of $\kappa$ because it captures the correct mean and the standard deviation of the distribution no matter what the underlying distribution is.

@@

If our model can fully describe the birefringence effect, the true value of $\kappa$ should be a single value, and the Gaussian distribution should have a standard deviation of $0$.
If the standard deviation obtained is not $0$, it means our model cannot fully describe the birefringence effect.
It could be due to our model does not fully capture the physics of birefringence, or there is a systematic error in the data.

@@

@@

\begin{figure*}[ht]
    \script{violin_kappa.py}
    \includegraphics[width=\textwidth]{figures/violin_kappa.pdf}
    \caption{
        The violin plot shows the posterior of $\kappa$ for 69 events in this study.
        Each violin represents a different event.
        The color of the violins represents the quotient of the median and standard deviation of the posterior.
        The blue horizontal solid line represents the median of the restricted posterior of $\kappa$, and the blue dashed lines enclose the $90\%$ confidence interval of the restricted posterior of $\kappa$.
        The pink horizontal solid line represents $\kappa=0$.
    }
    \label{fig:violin_kappa}
\end{figure*}

@@

\begin{figure}[h]
    \script{corner_Gaussian.py}
    \includegraphics[width=\columnwidth]{figures/corner_Gaussian.pdf}
    \caption{
        The posterior of $\mu$ and $\sigma$ of the population distribution of $\kappa$.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        The orange line represents the median of $\mu$ and $\sigma$.
        The plot shows that the population constraint on $\kappa$ is consistent with GR.
    }
    \label{fig:corner_Gaussian}
\end{figure}

@@

\begin{figure}[h]
    \script{posterior_kappa.py}
    \includegraphics[width=\columnwidth]{figures/posterior_kappa.pdf}
    \caption{
        The generic and restricted posterior of $\kappa$.
        The blue solid line shows the restricted posterior of $\kappa$, and the blue dashed lines enclose the $90\%$ confidence interval.
        The orange solid line shows the generic posterior of $\kappa$, and the orange dashed lines enclose the $90\%$ confidence interval.
        The pink vertical solid line represents $\kappa=0$.
    }
    \label{fig:posterior_kappa}
\end{figure}

@@

\begin{figure}[h]
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE result with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE results done by us with both frequency independent and dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE result from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency-independent birefringence creates a degeneracy between $\kappa$ and $\iota$, while frequency-dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

>>>>

\begin{figure}
    \script{birefringence.py}
    \includegraphics[width=\columnwidth]{figures/birefringence.pdf}
    \caption{
        The observed Fourier amplitude of two polarizations of a simulated GW signal with and without birefringence.
        The amplitudes, as GR predicts, are shown in dotted lines with blue and orange colors representing the left-handed and right-handed polarizations.
        The amplitudes with birefringence are shown in solid lines with blue and orange colors representing the left-handed and right-handed polarizations.
        This plot shows the effect of amplitude birefringence with frequency dependence.
        \mi{@Thomas can you add a panel with the time-domain waveforms (maybe just for L or R)?}
    }
    \label{fig:birefringence}
\end{figure}

@@

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby}, that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline waveform.
We depart from the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3} \mi{@Thomas cite actual zenodo link}, with \textsc{IMRPhenomXPHM} as the reference waveform.
We apply a distance prior uniform in comoving volume \mi{@Thomas is this right?}, and set the prior on $\kappa$ to a uniform distribution between $-1$ and $1$.

\mi{@Thomas list here any special considerations for individual events}

@@

Since birefringence is a property of spacetime and we are targeting a propagation effect not intrinsic to any source, we should expect $\kappa$ to take the same value for all signals, whether it vanishes or not.
\mi{caveats?}
Under this assumption, the posterior on $\kappa$ inferred collectively from all events with a uniform prior is simply the product of the individual posteriors,%
\footnote{Had the $\kappa$ prior been nonuniform, then we would need to multiply the individual \emph{likelihoods}, not to the posteriors.}
\begin{equation}
    p(\kappa \mid \{d_i\})\propto \prod_{i}p(\kappa \mid d_i)\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event.
We use this expression to obtain the primary constraint presented in this work.

@@

% Bayesian Hierarchical Modeling
In order to further characterize our set of measurements, and identify potential outliers, we carry out an additional collective analysis that allows $\kappa$ to take different values for each event \cite{Zimmerman:2019wzo,Isi:2022cii}.
To do this, we apply hierarchical Bayesian inference \cite{Loredo:2004nn} to model the distribution of $\kappa$'s consistent with the observed data:
we posit that there is a specific value of the parameter, $\kappa_i$, associated to each event, drawn from some unknown distribution of underlying values; from the imperfect measurements of $\kappa_i$ for each event, we may reconstruct the underlying distribution.
If we are interested in constraining the first two moments of the distribution, it is convenient to parametrize the $\kappa_i$'s as drawn from a Gaussian with unknown mean $\mu$ and variance $\sigma^2$, i.e., $\kappa_i \sim \mathcal{N}(\mu, \sigma^2)$ \cite{Isi:2019asy}, and measure those hyperparameters from the collection of observed data.

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements are statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

@@

@@

\begin{figure*}
    \script{violin_kappa.py}
    \includegraphics[width=\textwidth]{figures/violin_kappa.pdf}
    \caption{
        The violin plot shows the posterior of $\kappa$ for 69 events in this study.
        Each violin represents a different event.
        The color of the violins represents the quotient of the median and standard deviation of the posterior.
        The blue horizontal solid line represents the median of the restricted posterior of $\kappa$, and the blue dashed lines enclose the $90\%$ confidence interval of the restricted posterior of $\kappa$.
        The pink horizontal solid line represents $\kappa=0$.
    }
    \label{fig:violin_kappa}
\end{figure*}

@@

@@

\begin{figure}
    \script{corner_Gaussian.py}
    \includegraphics[width=\columnwidth]{figures/corner_Gaussian.pdf}
    \caption{
        The posterior of $\mu$ and $\sigma$ of the population distribution of $\kappa$.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        The orange line represents the median of $\mu$ and $\sigma$.
        The plot shows that the population constraint on $\kappa$ is consistent with GR.
    }
    \label{fig:corner_Gaussian}
\end{figure}

@@

\begin{figure}
    \script{posterior_kappa.py}
    \includegraphics[width=\columnwidth]{figures/posterior_kappa.pdf}
    \caption{
        The generic and restricted posterior of $\kappa$.
        The blue solid line shows the restricted posterior of $\kappa$, and the blue dashed lines enclose the $90\%$ confidence interval.
        The orange solid line shows the generic posterior of $\kappa$, and the orange dashed lines enclose the $90\%$ confidence interval.
        The pink vertical solid line represents $\kappa=0$.
    }
    \label{fig:posterior_kappa}
\end{figure}

@@

\begin{figure}
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        Colors in the plot represent the PE result with GR done by LVK without cosmological reweighing \citep{GWTC-2.1, GWTC-3}, the PE results done by us with both frequency independent and dependent birefringence respectively.
        2D contour plot shows $39.35\%$ and $90\%$ confidence level.
        Note that there is no posterior of $\kappa$ for the PE result from LVK, as the LVK PE is based on GR, which does not suggest GW amplitude birefringence.
        This plot shows that frequency-independent birefringence creates a degeneracy between $\kappa$ and $\iota$, while frequency-dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}
